High-Dimensional Regularized Discriminant Analysis

### 2016-12-21 reports/paper-hdrda.Rnw
Update paper-hdrda.Rnw
Phil's updates

We develop a new and computationally efficient implementation of
\citeauthor{Friedman:1989tm}'s popular regularized discriminant analysis (RDA)
classifier in the small-sample, high-dimensional setting. Our proposed
high-dimensional RDA (HDRDA) classifier modifies \citeauthor{Friedman:1989tm}'s
original formulation to achieve interpretability previously lacking. We are able
to interpret the biased covariance matrix estimator employed in the HDRDA
classifier as one that is estimated from all training observations regardless of
class, solidifying the ``borrowing'' operation \citeauthor{Friedman:1989tm}'s
RDA is claimed to have. Moreover, our formulation also differs from
\citeauthor{Friedman:1989tm}'s in that we show our classifier is equivalent to
one in a reduced feature space with dimension approximately equal to the
training sample size. As a result, the matrix operations used in the HDRDA
classifier are computationally linear in the number of features, making HDRDA
well-suited for high-dimensional classification in practice. Timing comparisons
between our HDRDA implementation in the {\tt sparsediscrim} R package and the
standard RDA formulation in the {\tt klaR} R package demonstrate that as the
number of features increases, the computational runtime of HDRDA is drastically
smaller than RDA. We also demonstrate that HDRDA is often superior to several
sparse and regularized classifiers in terms of classification accuracy with
three artificial and six real high-dimensional data sets.

@@

Here, we present the high-dimensional \emph{RDA} (\emph{HDRDA}) classifier,
which is intended for the $p > N$ case and builds on the classic \emph{RDA}
classifier proposed by \cite{Friedman:1989tm}. We reparameterize the \emph{RDA}
classifier similar to that of \cite*{Hastie:2008dt} and \cite{Halbe:2007ea} and
employ a biased covariance-matrix estimator that partially pools the individual
sample covariance matrices from the \emph{QDA} classifier with the pooled sample
covariance matrix from the \emph{LDA} classifier. We then shrink the resulting
covariance-matrix estimator towards a scaled identity matrix to ensure positive
definiteness. We show that the pooling parameter in the \emph{HDRDA} classifier
determines the contribution of each training observation to the estimation of
each class covariance matrix, enabling interpretability that has been previously
lacking with the \emph{RDA} classifier \citep{Bensmail:1996fw}. Our
parameterization differs from \cite*{Hastie:2008dt} and \cite{Halbe:2007ea} in
that our formulation allows the flexibility of various covariance-matrix
estimators proposed in the literature, including a variety of ridge-like
estimators such as the one proposed by \cite{Srivastava:2007ww}.

@@

We provide an efficient algorithm along with pseudocode to estimate the
\emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Timing comparisons between our \emph{HDRDA} implementation in
the {\tt sparsediscrim} R package available on CRAN and the standard \emph{RDA}
formulation in the {\tt klaR} R package demonstrate that as the number of
features increases, the computational runtime of the \emph{HDRDA} classifier is
drastically smaller than \emph{RDA}. In fact, when $p = 5000$, we show that the
\emph{HDRDA} classifier is \Sexpr{bootstrap_summary_p5000$avg_ratio} times
faster on average than the \emph{RDA} classifier. In this scenario, the
\emph{HDRDA} classifier's training and model selection requires 3 seconds on
average, while that of the \emph{RDA} classifier requires 25 minutes on average.

@@

Perhaps the most well-known of the ridge-like estimators is from
\cite{Friedman:1989tm}, who proposed regularized discriminant analysis (RDA) by
incorporating a weighted average of the sample covariance matrix $\widehat{ \bm
  \Sigma}_k$ and the pooled sample covariance matrix $\widehat{\bm \Sigma}$ to
estimate $\bm \Sigma_k$ with
\begin{linenomath*}
\begin{align}
    \widehat{\bm \Sigma}_k(\lambda) = n_k^{-1}(\lambda) \bm S_k(\lambda) \quad
    (k = 1, \ldots, K),\label{eq:sig-lambda}
\end{align}
\end{linenomath*}
where $\lambda \in [0, 1]$, $n_k(\lambda) = (1 - \lambda) n_k + \lambda N$, $
\bm S_k = n_k \widehat{\bm \Sigma}_k$, $\bm S = \sum_{k=1}^K \bm S_k$, and $ \bm
S_k(\lambda) = (1 - \lambda) \bm S_k + \lambda \bm S$. If $\lambda = 0$,
\eqref{eq:sig-lambda} is equal to $\widehat{\bm \Sigma}_k$ in \eqref{eq:qda}.
Alternatively, if $\lambda = 1$, then \eqref{eq:sig-lambda} reduces to
\eqref{eq:pooled-cov}. To further improve the estimation of $\bm \Sigma_k$ and
to stabilize the inverse of \eqref{eq:sig-lambda}, \cite{Friedman:1989tm}
proposed the biased covariance-matrix estimator
\begin{linenomath*}
\begin{align}
    \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
\end{align}
\end{linenomath*}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. Finally, \cite{Friedman:1989tm}
established the RDA classifier by substituting \eqref{eq:sig-rda} into
\eqref{eq:qda}.

Our proposed \emph{HDRDA} classifier resembles \eqref{eq:sig-rda} but yields an
interpretation of the pooling parameter $\lambda$ as discussed above. Moreover,
as we discuss below, our formulation of the \emph{HDRDA} classifier lends itself
to rapid estimation in high-dimensional classification.

@@

Thus, we can avoid the time-consuming inverses and determinants of $p \times p$
covariance matrices in \eqref{eq:hdrda} and instead calculate these same
operations on $\bm W_k \in \mathbb{R}_{q \times q}$ in
\eqref{eq:hdrda-decomposed}. The substantial computational improvements arise
because our proposed classifier in \eqref{eq:hdrda} is invariant to the term
$\bm U_2$, thus yielding an equivalent classifier in \eqref{eq:hdrda-decomposed}
with a substantial reduction in computational complexity. Here, we demonstrate
that the computational efficiency in calculating the inverse and determinant of
$\bm W_k$ can be further improved via standard matrix operations when we show
that the inverses and determinants of $\bm W_k$ can be performed on matrices of
size $n_k \times n_k$.

@@

The application of the Sherman-Woodbury formula requires that $\bm \Gamma_k$ be
nonsingular. This assumption is not met when $(\lambda, \gamma) = (0, 0)$
because $\bm \Gamma_k = \bm 0_q$. Here, we use the formulation in
\eqref{eq:hdrda-decomposed} instead. Also, notice that if $\alpha_k$ is constant
across the $K$ classes, then $\bm \Gamma_k$ in \eqref{eq:Gamma_k} is independent
of $k$. Consequently, $|\bm \Gamma_k|$ is constant across the $K$ classes and
need not be calculated in \eqref{eq:hdrda-decomposed}.

@@

The expressions given in Proposition \ref{proposition:hdrda-W_k} also expedite
the selection of $\lambda$ and $\gamma$ via cross-validation because the most
time-consuming matrix operation involved in computing $\bm W_k^{-1}$ and $|\bm
W_k|$ is $\bm X_k \bm U_1 \in \mathbb{R}_{n_k \times q}$, which is independent
of $\lambda$ and $\gamma$. The subsequent operations in calculating $\bm
W_k^{-1}$ and $|\bm W_k|$ can be simply updated for different pairs of $\lambda$
and $\gamma$ without repeating the costly computations. When calculating the
discriminant scores of each test observation in $\bm X_{test}$, rather than
calculating each score individually, we can more efficiently calculate the
quadratic forms via a single matrix operation. That is, rather than calculating
$(\bm x - \xbar_k)^{T} \bm U_1 \bm W_k^{-1} \bm U_1^{T} (\bm x - \xbar_k)$
individually for each row $\bm x$ of $\bm X_{test}$, we can calculate $(\bm
X_{test} - \xbar_k \bm 1_k')' \bm U_1 \bm W_k^{-1} \bm U_1^{T} (\bm X_{test} -
\xbar_k \bm 1_k')$. The diagonal elements of the resulting matrix contain the
individual quadratic form of each test observation, $\bm x_t$.

@@

In this section, we demonstrate that the computational performance of the
training and model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as the number of features increases. To
compare the two classifiers, we generate 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimate the $\lambda$ and
$\gamma$ parameters for both classifiers using a grid of 5 equidistant candidate
values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$, $k = 1,
\ldots, K$, in the \emph{HDRDA} classifier to resemble
\citeauthor{Friedman:1989tm}'s \emph{RDA} parameterization. At each pair of
$\lambda$ and $\gamma$, we compute the 10-fold cross-validation error rate
\citep{Hastie:2008dt}. Then, we select the optimal model that minimizes the 10-fold
cross-validation error rate.

Here, our goal is to compare the computational runtime of the model selection
for both classifiers as a function of the number of features $p$. We compare the
runtime of both classifiers by increasing the number of features from $p = 500$
to $p = 5000$ in increments of 500. We generate 100 data sets for each value of
$p$ and compute the training and model selection runtime of both
classifiers. Our timing comparisons are based on our \emph{HDRDA} implementation
in the {\tt sparsediscrim} R package and the standard \emph{RDA} implementation
in the {\tt klaR} R package. All timing comparisons were conducted on an Amazon
Elastic Compute Cloud (EC2) {\tt c4.4xlarge} instance using version 3.3.1 of the
open-source statistical software {\tt R}. Our timing comparisons can be
reproduced with the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

@@

In Figure \ref{fig:timing-results}, we plot the runtime of the model selections
for both the \emph{HDRDA} and \emph{RDA} classifiers as a function of the number
of features $p$. In the left panel of Figure \ref{fig:timing-results}, we
observe that the \emph{HDRDA} classifier is substantially faster than the
\emph{RDA} classifier as $p$ increases. Figure \ref{fig:timing-results} suggests
that the usage of a matrix inverse and determinant in the {\tt klaR} R package's
discriminant function yielded training and model-selection timings that exceeded
linear growth in $p$. Because our implementation in the \emph{HDRDA} classifier
removes the inverse and determinants from the classifier, the \emph{HDRDA}
classifier is computationally more efficient than the \emph{RDA} classifier,
especially as $p$ increases. We contrast our implementation to the the poor
computational performance of the in the {\tt klaR} R package, which took
approximately 25 minutes on average to train the \emph{RDA} classifier on a data
set having 5,000 features. To quantify how much beyond linear growth in $p$, we
fit a regression line quadratic in $p$ to the \emph{RDA} runtimes and plotted
the fit in Figure \ref{fig:timing-results}. We attempted to fit a cubic
polynomial in $p$ to the \emph{RDA} timings, but the cubic term was
insignificant with significant quadratic and linear terms. We also fit two
polynomials to the \emph{RDA} timings with orders 2.807 and 2.376, which
correspond to the asymptotic computational runtime of two well-known algorithms
for computing a matrix inverse: the Strassen and Coppersmith-Winograd
algorithms, respectively \footnote{TODO: Citations needed}. In both cases, the
polynomials were statistically significiant at the 0.05 level. We displayed the
quadratic fit in Figure \ref{fig:timing-results} because it fit the timings best
for $p = 500$ to $p = 2000$. The fits for $2000 \le p \le 5000$ were similar.

In the left panel of Figure \ref{fig:timing-results}, we also fit a linear
regression model that was linear in $p$ to the \emph{HDRDA} runtimes. In the
right panel, we repeated the same scatterplot and linear fit with the timings
restricted to the observed range of \emph{HDRDA} timings. The outliers in the
\emph{HDRDA} timings were primarily due to the inherent computing variability in
our {\tt c4.4xlarge} instance.

Clearly, the model training and selection employed in \emph{HDRDA} is
substantially faster than that of \emph{RDA}. To investigate the relative timing
comparisons between the two classifiers, we consider the ratio of mean timings
of the \emph{RDA} classifier to the \emph{HDRDA} classifier for each value of
$p$. We employed nonparametric bootstrapping\footnote{TODO: Citation needed} to
estimate the mean ratio along with 95\% confidence intervals. In Figure
\ref{fig:timing-comparison-bootstrap}, the bootstrap sampling distributions for
the ratio of mean timings are given. First, we observe that the mean relative
timings increases as $p$ increases. For smaller dimensions, the relative
difference in computing is sizable with the average ratio of the mean timings
equal to \Sexpr{bootstrap_summary_p500$avg_ratio} for $p = 500$ and a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p500$lower_ci},
\Sexpr{bootstrap_summary_p500$upper_ci}). Furthermore, the ratio of mean
computing times suggests that the \emph{RDA} classifier is impractical for
higher dimensions. For instance, when $p = 5000$ the ratio of mean computing
times increases to \Sexpr{bootstrap_summary_p5000$avg_ratio} with a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p5000$lower_ci},
\Sexpr{bootstrap_summary_p5000$upper_ci}).

@@

For the \emph{HDRDA} classifier in \eqref{eq:hdrda-decomposed}, we examined the
classification performance of two models. For the first \emph{HDRDA} model, we
set $\alpha_k = 1$, $k = 1, \ldots, K$, so that the covariance-matrix estimator
\eqref{eq:hdrda-cov} resembled \eqref{eq:ridge-estimator}. We estimated
$\lambda$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Similarly, we estimated $\gamma$ from a grid consisting of the
values $10^{-1}, \ldots, 10^4$, and $10^5$. We selected optimal estimates of
$\lambda$ and $\gamma$ using $V$-fold cross-validation with $V = 10$. For the
second model, we set $\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization, and we estimated both $\lambda$
and $\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively.

@@

Although the design is indeed standard, the simulation configuration lacks
artifacts observed commonly in real data, such as skewness and extreme
outliers. As a result, we wished to investigate the effect of outliers on the
high-dimensional classifiers. To accomplish this goal, we generalized the
block-diagonal simulation configuration by sampling from a $p$-dimensional
multivariate contaminated normal distribution.  Denoting the PDF of the
$p$-dimensional multivariate normal distribution by $N_p(\bm x | \bm \mu, \bm
\Sigma)$, we write the PDF of the $k$th class as
\begin{linenomath*}
\begin{align}
  p(\bm x | \omega_k) = (1 - \epsilon) N_p(\bm x | \bm \mu_k, \bm \Sigma_k) + \epsilon N_p(\bm x | \bm \mu_k, \eta \bm \Sigma_k),\label{eq:contaminated-pdf}
\end{align}
\end{linenomath*}
where $\epsilon \in [0, 1]$ is the probability that an observation is
contaminated (i.e., drawn from a distribution with larger variance) and $\eta >
1$ scales the covariance matrix $\bm \Sigma_k$ to increase the extremity of
outliers. For $\epsilon = 0$, we have the benchmark block-diagonal simulation
design from \cite{Guo:2007te}. As $\epsilon$ is increased, the average number of
outliers is increased. In our simulation, we let $\eta = 100$ and considered the
values of $\epsilon = 0$, $0.05$, $\ldots$, $0.50$.

We generated $K=3$ populations from \eqref{eq:contaminated-pdf} with $\bm
\Sigma_k$ given in \eqref{eq:block-sigma} and set the mean vector of class 1 to
$\bm \mu_1 = \bm 0_p$. Next, comparable to \cite{Guo:2007te}, the first 100
features of $\bm \mu_2$ were set to 1/2, while the rest were set to 0, i.e.,
$\bm \mu_2 = (\underbrace{1/2, \ldots, 1/2}_{100}, \underbrace{0, \ldots, 0}_{p
  - 100})$. For simplicity, we defined $\bm \mu_3 = -\bm \mu_2$. The three
populations differed in their mean vectors in the first 100 features
corresponding to the first block. No difference in the means occurred in the
remaining blocks.

@@

In Figure \ref{fig:sim-results}, we observe each classifier's average
classification error rates for the values of $\epsilon$ and $p$. Unsurprisingly,
the average error rate increased for each classifier as the contamination
probability $\epsilon$ increased regardless of the value of $p$. Sensitivity to
the presence of outliers was most apparent for the Pang, Tong, and Witten
classifiers. For smaller dimensions, the random-forest and \emph{HDRDA}
classifiers tended to outperform the remaining classifiers with the random
forest performing best. As the feature dimension increased with $p \ge 300$,
both \emph{HDRDA} classifiers outperformed all other classifiers, suggesting
that their inherent dimension reduction better captured the classificatory
information in the small training samples, even in the presence of outliers.

@@

In Figure \ref{fig:sim-results-by-p} we again examine the simulation results as
a function of the number of features $p$ for a subset of the values of
$\epsilon$. This set of plots allows us to investigate the effect of feature
dimensionality on classification performance. When no outliers were present
(i.e., $\epsilon=0$), the random-forest classifier was outperformed by all other
classifiers. Furthermore, the \emph{HDRDA} classifiers were superior in terms of
average error rate in this setting. As $p$ increased, an elevation in average
error rate was expected for all classifiers, but the increase was not observed
to be substantial.

@@

\cite{Nakayama:2007fl} acquired 105 gene-expression samples of 10 types of
soft-tissue tumors through an oligonucleotide microarray, including 16 samples
of synovial sarcoma (SS), 19 samples of myxoid/round cell liposarcoma (MLS), 3
samples of lipoma, 3 samples of well-differentiated liposarcoma (WDLS), 15
samples of dedifferentiated liposarcoma (DDLS), 15 samples of myxofibrosarcoma
(MFS), 6 samples of leiomyosarcoma (LMS), 3 samples of malignant nerve sheathe
tumor (MPNST), 4 samples of fibrosarcoma (FS), and 21 samples of malignant
fibrous histiocytoma (MFH). \cite{Nakayama:2007fl} determined from their data
that these 10 types fell into 4 broader groups: (1) SS; (2) MLS; (3) Lipoma,
WDLS, and part of DDLS; (4) Spindle cell and pleomorophic sarcomas including
DDLS, MFS, LMS, MPNST, FS, and MFH. Following \cite{Witten:2011kc}, we restrict
our analysis to the five tumor types having at least 15 observations.

>>>>

We develop a new and computationally efficient regularized discriminant analysis (RDA)
classifier for the small-sample, high-dimensional setting.  We are able
to interpret the biased covariance matrix estimator employed in the HDRDA
classifier as one that is estimated from all training observations regardless of
class. Moreover, our classifier also differs from
\citeauthor{Friedman:1989tm}'s in that we show our classifier is equivalent to
a classifier in a reduced feature space with dimension approximately equal to the
training sample size. As a result, the matrix operations used in the HDRDA
classifier are computationally linear in the number of features, making HDRDA
well-suited for high-dimensional classification in practice. We demonstrate that HDRDA is often superior to several
sparse and regularized classifiers in terms of classification accuracy with
three artificial and six real high-dimensional data sets.  Also, timing comparisons
between our HDRDA implementation in the {\tt sparsediscrim} R package and the
standard RDA formulation in the {\tt klaR} R package demonstrate that as the
number of features increases, the computational runtime of HDRDA is drastically
smaller than RDA.

@@

Here, we present the high-dimensional \emph{RDA} (\emph{HDRDA}) classifier,
which is intended for the case when $p > N$ . We reparameterize the \emph{RDA}
classifier similar to that of \cite*{Hastie:2008dt} and \cite{Halbe:2007ea} and
employ a biased covariance-matrix estimator that partially pools the individual
sample covariance matrices from the \emph{QDA} classifier with the pooled sample
covariance matrix from the \emph{LDA} classifier. We then shrink the resulting
covariance-matrix estimator towards a scaled identity matrix to ensure positive
definiteness. We show that the pooling parameter in the \emph{HDRDA} classifier
determines the contribution of each training observation to the estimation of
each class covariance matrix, enabling interpretability that has been previously
lacking with the \emph{RDA} classifier \citep{Bensmail:1996fw}. Our
parameterization differs from \cite*{Hastie:2008dt} and \cite{Halbe:2007ea} in
that our formulation allows the flexibility of various covariance-matrix
estimators proposed in the literature, including a variety of ridge-like
estimators such as the one proposed by \cite{Srivastava:2007ww}.

@@

We also provide an efficient algorithm along with pseudocode to estimate the
\emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Timing comparisons between our \emph{HDRDA} implementation in
the {\tt sparsediscrim} R package available on CRAN and the standard \emph{RDA}
formulation in the {\tt klaR} R package demonstrate that as the number of
features increases, the computational runtime of the \emph{HDRDA} classifier is
drastically smaller than \emph{RDA}. In fact, when $p = 5000$, we show that the
\emph{HDRDA} classifier is \Sexpr{bootstrap_summary_p5000$avg_ratio} times
faster on average than the \emph{RDA} classifier. In this scenario, the
\emph{HDRDA} classifier's training and model selection requires 3 seconds on
average, while that of the \emph{RDA} classifier requires 25 minutes on average.

@@

\cite{Friedman:1989tm}
proposed the biased covariance-matrix estimator
\begin{linenomath*}
\begin{align}
    \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
\end{align}
\end{linenomath*}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. \cite{Friedman:1989tm}
established the RDA classifier by substituting \eqref{eq:sig-rda} into
\eqref{eq:qda}.

Our proposed \emph{HDRDA} classifier yields an
interpretation of the pooling parameter $\lambda$ as discussed above. Moreover,
as we discuss below, our formulation of the \emph{HDRDA} classifier lends itself
to rapid estimation in high-dimensional classification.

@@

Using Theorem 1, we can avoid the time-consuming inverses and determinants of $p \times p$
covariance matrices in \eqref{eq:hdrda} and instead calculate these same
operations on $\bm W_k \in \mathbb{R}_{q \times q}$ in
\eqref{eq:hdrda-decomposed}. The substantial computational improvements arise
because our proposed classifier in \eqref{eq:hdrda} is invariant to the term
$\bm U_2$, thus yielding an equivalent classifier in \eqref{eq:hdrda-decomposed}
with a substantial reduction in computational complexity. Here, we demonstrate
that the computational efficiency in calculating the inverse and determinant of
$\bm W_k$ can be further improved via standard matrix operations when we show
that the inverses and determinants of $\bm W_k$ can be performed on matrices of
size $n_k \times n_k$.

@@

Also, notice that if $\alpha_k$ is constant
across the $K$ classes, then $\bm \Gamma_k$ in \eqref{eq:Gamma_k} is independent
of $k$. Consequently, $|\bm \Gamma_k|$ is constant across the $K$ classes and
need not be calculated in \eqref{eq:hdrda-decomposed}.

@@

The expressions given in Proposition \ref{proposition:hdrda-W_k} also expedite
the selection of $\lambda$ and $\gamma$ via cross-validation because the most
time-consuming matrix operation involved in computing $\bm W_k^{-1}$ and $|\bm
W_k|$ is $\bm X_k \bm U_1 \in \mathbb{R}_{n_k \times q}$, which is independent
of $\lambda$ and $\gamma$. The subsequent operations in calculating $\bm
W_k^{-1}$ and $|\bm W_k|$ can be simply updated for different pairs of $\lambda$
and $\gamma$ without repeating the costly computations. Also, rather than calculating
$(\bm x - \xbar_k)^{T} \bm U_1 \bm W_k^{-1} \bm U_1^{T} (\bm x - \xbar_k)$
individually for each row $\bm x$ of $\bm X_{test}$, we can calculate $(\bm
X_{test} - \xbar_k \bm 1_k')' \bm U_1 \bm W_k^{-1} \bm U_1^{T} (\bm X_{test} -
\xbar_k \bm 1_k')$. The diagonal elements of the resulting matrix contain the
individual quadratic form of each test observation, $\bm x_t$.

@@

In this section, we demonstrate that the computational performance of the
training and model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as $p$ increases. To
compare the two classifiers, we generated 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimated the parameters $\lambda$ and
$\gamma$ for both classifiers using a grid of 5 equidistant candidate
values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$, $k = 1,
\ldots, K$, in the \emph{HDRDA} classifier. At each pair of
$\lambda$ and $\gamma$, we computed the 10-fold cross-validation error rate
\citep{Hastie:2008dt}. Then, we selected the optimal model that minimized the 10-fold
cross-validation error rate.

Here, our goal was to compare the computational runtime of the model selection
for both classifiers as a function of $p$. We compared the
runtime of both classifiers by increasing the number of features from $p = 500$
to $p = 5000$ in increments of 500. We generated 100 data sets for each value of
$p$ and computed the training and model selection runtime of both
classifiers. Our timing comparisons are based on our \emph{HDRDA} implementation
in the {\tt sparsediscrim} R package and the standard \emph{RDA} implementation
in the {\tt klaR} R package. All timing comparisons were conducted on an Amazon
Elastic Compute Cloud (EC2) {\tt c4.4xlarge} instance using version 3.3.1 of the
open-source statistical software {\tt R}. Our timing comparisons can be
reproduced with the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

@@

In Figure \ref{fig:timing-results}, we plotted the runtime of the model selections
for both the \emph{HDRDA} and \emph{RDA} classifiers as a function of $p$. In the left panel of Figure \ref{fig:timing-results}, we
observe that the \emph{HDRDA} classifier is substantially faster than the
\emph{RDA} classifier as $p$ increases. Figure \ref{fig:timing-results} suggests
that the usage of a matrix inverse and determinant in the {\tt klaR} R package's
discriminant function yielded training and model-selection timings that exceeded
linear growth in $p$. Because our implementation in the \emph{HDRDA} classifier
removes the inverse and determinants from the classifier, the \emph{HDRDA}
classifier is computationally more efficient than the \emph{RDA} classifier,
especially as $p$ increases. We contrast our implementation to the the poor
computational performance of the in the {\tt klaR} R package, which took
approximately 25 minutes on average to train the \emph{RDA} classifier on a data
set having 5,000 features. To quantify how much beyond linear growth in $p$, we
fit a regression line quadratic in $p$ to the \emph{RDA} runtimes and plotted
the fit in Figure \ref{fig:timing-results}. We displayed the
quadratic fit in Figure \ref{fig:timing-results} because it fit the timings best
for $p = 500$ to $p = 2000$. The fits for $2000 \le p \le 5000$ were similar.

Clearly, the model training and selection employed in \emph{HDRDA} is
substantially faster than that of \emph{RDA}. To investigate the relative timing
comparisons between the two classifiers, we considered the ratio of mean timings
of the \emph{RDA} classifier to the \emph{HDRDA} classifier for each value of
$p$. We employed nonparametric bootstrapping\footnote{TODO: Citation needed} to
estimate the mean ratio along with 95\% confidence intervals. In Figure
\ref{fig:timing-comparison-bootstrap}, the bootstrap sampling distributions for
the ratio of mean timings are given. First, we observe that the mean relative
timings increased as $p$ increased. For smaller dimensions, the relative
difference in computing was sizable with the average ratio of the mean timings
equal to \Sexpr{bootstrap_summary_p500$avg_ratio} for $p = 500$ and a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p500$lower_ci},
\Sexpr{bootstrap_summary_p500$upper_ci}). Furthermore, the ratio of mean
computing times suggested that the \emph{RDA} classifier is impractical for
higher dimensions. For instance, when $p = 5000$ the ratio of mean computing
times increased to \Sexpr{bootstrap_summary_p5000$avg_ratio} with a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p5000$lower_ci},
\Sexpr{bootstrap_summary_p5000$upper_ci}).

@@

For the \emph{HDRDA} classifier in \eqref{eq:hdrda-decomposed}, we examined the
classification performance of two models. For the first \emph{HDRDA} model, we
set $\alpha_k = 1$, $k = 1, \ldots, K$, so that the covariance-matrix estimator
\eqref{eq:hdrda-cov} resembled \eqref{eq:ridge-estimator}. We estimated
$\lambda$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Similarly, we estimated $\gamma$ from a grid consisting of the
values $10^{-1}, \ldots, 10^4$, and $10^5$. We selected optimal estimates of
$\lambda$ and $\gamma$ using $10$-fold cross-validation. For the
second model, we set $\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization, and we estimated both $\lambda$
and $\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively.

@@

Although the design is indeed standard, the simulation configuration lacks
artifacts commonly observed in real data, such as skewness and extreme
outliers. As a result, we wished to investigate the effect of outliers on the
high-dimensional classifiers. To accomplish this goal, we generalized the
block-diagonal simulation configuration by sampling from a $p$-dimensional
multivariate contaminated normal distribution.  Denoting the PDF of the
$p$-dimensional multivariate normal distribution by $N_p(\bm x | \bm \mu, \bm
\Sigma)$, we write the PDF of the $k$th class as
\begin{linenomath*}
\begin{align}
  p(\bm x | \omega_k) = (1 - \epsilon) N_p(\bm x | \bm \mu_k, \bm \Sigma_k) + \epsilon N_p(\bm x | \bm \mu_k, \eta \bm \Sigma_k),\label{eq:contaminated-pdf}
\end{align}
\end{linenomath*}
where $\epsilon \in [0, 1]$ is the probability that an observation is
contaminated (i.e., drawn from a distribution with larger variance) and $\eta >
1$ scales the covariance matrix $\bm \Sigma_k$ to increase the extremity of
outliers. For $\epsilon = 0$, we have the benchmark block-diagonal simulation
design from \cite{Guo:2007te}. As $\epsilon$ is increased, the average number of
outliers is increased. In our simulation, we let $\eta = 100$ and considered the
values of $\epsilon = 0$, $0.05$, $\ldots$, $0.50$.

We generated $K=3$ populations from \eqref{eq:contaminated-pdf} with $\bm
\Sigma_k$ given in \eqref{eq:block-sigma} and set the mean vector of class 1 to
$\bm \mu_1 = \bm 0_p$. Next, comparable to \cite{Guo:2007te}, the first 100
features of $\bm \mu_2$ were set to 1/2, while the rest were set to 0, i.e.,
$\bm \mu_2 = (\underbrace{1/2, \ldots, 1/2}_{100}, \underbrace{0, \ldots, 0}_{p
  - 100})$. For simplicity, we defined $\bm \mu_3 = -\bm \mu_2$. The three
populations differed in their mean vectors in the first 100 features
corresponding to the first block, and no difference in the means occurred in the
remaining blocks.

@@

In Figure \ref{fig:sim-results}, we observed each classifier's average
classification error rates for the values of $\epsilon$ and $p$. Unsurprisingly,
the average error rate increased for each classifier as the contamination
probability $\epsilon$ increased regardless of the value of $p$. Sensitivity to
the presence of outliers was most apparent for the Pang, Tong, and Witten
classifiers. For smaller dimensions, the random-forest and \emph{HDRDA}
classifiers tended to outperform the remaining classifiers with the random
forest performing best. As the feature dimension increased with $p \ge 300$,
both \emph{HDRDA} classifiers outperformed all other classifiers, suggesting
that their inherent dimension reduction better captured the classificatory
information in the small training samples, even in the presence of outliers.

@@

In Figure \ref{fig:sim-results-by-p} we again examine the simulation results as
a function of $p$ for a subset of the values of
$\epsilon$. This set of plots allows us to investigate the effect of feature
dimensionality on classification performance. When no outliers were present
(i.e., $\epsilon=0$), the random-forest classifier was outperformed by all other
classifiers. Furthermore, the \emph{HDRDA} classifiers were superior in terms of
average error rate in this setting. As $p$ increased, an elevation in average
error rate was expected for all classifiers, but the increase was not observed
to be substantial.

@@

\cite{Nakayama:2007fl} acquired 105 gene-expression samples of 10 types of
soft-tissue tumors through an oligonucleotide microarray, including 16 samples
of synovial sarcoma (SS), 19 samples of myxoid/round cell liposarcoma (MLS), 3
samples of lipoma, 3 samples of well-differentiated liposarcoma (WDLS), 15
samples of dedifferentiated liposarcoma (DDLS), 15 samples of myxofibrosarcoma
(MFS), 6 samples of leiomyosarcoma (LMS), 3 samples of malignant nerve sheathe
tumor (MPNST), 4 samples of fibrosarcoma (FS), and 21 samples of malignant
fibrous histiocytoma (MFH). \cite{Nakayama:2007fl} determined from their data
that these 10 types fell into 4 broader groups: (1) SS; (2) MLS; (3) Lipoma,
WDLS, and part of DDLS; (4) Spindle cell and pleomorophic sarcomas including
DDLS, MFS, LMS, MPNST, FS, and MFH. Following \cite{Witten:2011kc}, we restricted
our analysis to the five tumor types having at least 15 observations.
